---
title: "Template Syntax"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Template Syntax}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette follows the structure of the Jinja [Template Designer Documentation](https://jinja.palletsprojects.com/templates/).

```{r setup}
library(jinjar)
```


## Synopsis



## Variables

When writing a template, we refer to variables that act as data placeholders.
We define their values when rendering the template.

Although we pass R objects to `render()`, it is helpful to understand that these are encoded as JSON objects before the template is rendered.

| R object        | JSON object      | Template example  |
|:----------------|:-----------------|:------------------|
| Length-1 vector | Scalar           | `{{ foo }}`       |
| Vector          | Array            | `{{ foo.1 }}`     |
| List            | Object           | `{{ foo.bar }}`   |
| Data frame      | Array of objects | `{{ foo.1.bar }}` |

You can use dot (`.`) notation to access data nested within a variable.
An array element is accessed by its numeric index (e.g. `foo.1`) and an object value is accessed by its key (e.g. `foo.bar`).

**Note:** In R, the dot is a valid character in an object name (e.g. `my.data`).
However, this causes ambiguity when accessing nested data values.
For this reason, each dot is replaced with an underscore when the data is encoded as JSON (e.g. `my.data` becomes `my_data`).

The double-brace syntax is used to print the value of the variable (e.g. `{{ foo }}`).
To use the variable in other contexts (e.g. control structures), then these braces are omitted (e.g. `{% for bar in foo %}`).

If a template variable has not been defined, then an error occurs.
However, you can use the `default(foo, bar)` function to specify a fallback value.

## Comments

## Whitespace Control

## Line Statements

## Control Structures

### For

### If

### Assignments


## Expressions

### Literals

### Math

### Comparisons

### Logic

### Other Operators


## Functions
