---
title: "Template Syntax"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Template Syntax}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette describes the template syntax supported by the jinjar R package, following the structure of the Jinja [Template Designer Documentation](https://jinja.palletsprojects.com/templates/).
It is designed to act as a reference when writing templates, because there are many subtle differences between the jinjar and Jinja syntaxes.
The underlying reason is that jinjar is powered by the fantastic [inja](https://github.com/pantor/inja) C++ library.

Before starting, let's create a few R objects for rendering example templates.

```{r setup}
library(jinjar)

navigation <- data.frame(
  caption = c("Home", "Blog"),
  href = c("index.html", "blog.html")
)
title <- "My Webpage"
```

```{r, echo=FALSE}
params <- list(
  navigation = navigation,
  title = title
)
```


## Synopsis

A jinjar template is simply a text file, and when rendered the output is also a text file (e.g. HTML, SQL, LaTeX).

A template contains **variables** and/or **expressions**, which get replaced with values when a template is rendered; and **tags**, which control the logic of the template.

Below is a minimal template that illustrates a few basics using the default jinjar configuration.
We will cover the details later in this document:

```{jinjar, jinjar_lang="html", data=params}
<!DOCTYPE html>
<html lang="en">
<head>
    <title>{{ title }}</title>
</head>
<body>
    <ul id="navigation">
    {% for item in navigation -%}
        <li><a href="{{ item.href }}">{{ item.caption }}</a></li>
    {% endfor -%}
    </ul>
    {# a comment #}
</body>
</html>
```

The following example shows the default configuration settings, but you can adjust the syntax configuration as desired using `engine_config()`.

There are a few kinds of delimiters. The default delimiters are configured as follows:

* `{% ... %}` for [Statements](#control-structures)
* `{{ ... }}` for [Expressions](#expressions) to print to the template output
* `{# ... #}` for [Comments](#comments) not included in the template output

[Line Statements](#line-statements) are also possible, though they don’t have default prefix characters.
To use them, set `line_statement` when creating the `engine_config()`.


## Variables {#variables}

When writing a template, we refer to variables that act as data placeholders.
We define their values when rendering the template.

Although we pass R objects to `render()`, it is helpful to understand that these are encoded as JSON objects before the template is rendered.

| R object        | JSON object      | Template example  |
|:----------------|:-----------------|:------------------|
| Length-1 vector | Scalar           | `{{ foo }}`       |
| Vector          | Array            | `{{ foo.1 }}`     |
| List            | Object           | `{{ foo.bar }}`   |
| Data frame      | Array of objects | `{{ foo.1.bar }}` |

You can use dot (`.`) notation to access data nested within a variable.
An array element is accessed by its numeric index (e.g. `foo.1`) and an object value is accessed by its key (e.g. `foo.bar`).

**Note:** In R, the dot is a valid character in an object name (e.g. `my.data`).
However, this causes ambiguity when accessing nested data values.
For this reason, each dot is replaced with an underscore when the data is encoded as JSON (e.g. `my.data` becomes `my_data`).

The double-brace syntax is used to print the value of the variable (e.g. `{{ foo }}`).
To use the variable in other contexts (e.g. control structures), then these braces are omitted (e.g. `{% for bar in foo %}`).

If a template variable has not been defined, then an error occurs.
However, you can use the `default(foo, bar)` function to specify a fallback value.

## Comments {#comments}

To comment-out some lines, preventing them from appearing in the rendered document, use the comment syntax (default: `{# ... #}`).
This is useful for debugging or documenting the template.

```
{# note: commented-out template because we no longer use this
    {% for user in users %}
        ...
    {% endfor %}
#}
```

## Whitespace Control {#whitespace}

## Line Statements {#line-statements}

## Control Structures {#control-structures}

### For

### If

### Assignments


## Expressions {#expressions}

### Literals

### Math

### Comparisons

### Logic

### Other Operators


## Functions
